---
title: "LC-MS of *U.saggitifolia* in positive ionization mode"
author: "Pablo Corella, Jefferson Pastuña"
date: "2023-06-11"
output:
  github_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE)
usethis::git_vaccinate()
```

# Workspace
## Charging necesary libraries

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}
library(notame)  
library(doParallel) 
library(magrittr)  
library(tidyverse) 
library(patchwork) 
library(dplyr)
library(here)
library(gplots)
```

## Creating a record

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}
ppath <-here()
init_log(log_file = paste0(ppath, '/Result/LC-MS_Positive/log_LC-MS_Positive.txt'))
```

## Importing data

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}
POS_data <- read_from_excel(file =  "Data/POS_feature_list.xlsx", sheet = 1,
                            corner_row = 4, corner_column = "E",
                            split_by = c("Column", "Ion mode"))
```

## Creating metaboset objects

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}
POS_modes <- construct_metabosets(exprs = POS_data$exprs,
                                  pheno_data = POS_data$pheno_data,
                                  feature_data = POS_data$feature_data,
                                  group_col = "Group")
```

## Extracting each mode in a single object

```{r}
POS_mode <- POS_modes$RP_POS
```

## Plotting raw data

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}
POS_raw_sambx <- plot_sample_boxplots(POS_mode, order_by = "Group")
POS_raw_pca <- plot_pca(POS_mode, center = T)
POS_raw_pca + POS_raw_sambx
```

# Preprocessing

## Changing features with 0 to NA.

```{r}
POS_mode <- mark_nas(POS_mode, value = 0)
```

## Flagging low detection rate features

Features which that were not detected in the 75% of the QC injections and 80% of sample groups will be flagged.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}
POS_mode <- flag_detection(POS_mode, qc_limit = 0.75, group_limit = 0.80)
```

## Drift correction 

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}
POS_dc <- dc_cubic_spline(POS_mode)
POS_corrected <- POS_dc$object
POS_corrected <- flag_quality(POS_corrected)
```

## Plotting corrected data

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}
POS_corr_sambx <- plot_sample_boxplots(POS_corrected, order_by = "Group")
POS_corr_pca <- plot_pca(POS_corrected, center = T) 
POS_corr_pca + POS_corr_sambx
```

## Feature clustering 

This step helps us reduce the number of features of the same molecule that were split due to ionization problems or unexpected behavior.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide', cache=TRUE}
POS_clusteredQC <- cluster_features(POS_corrected, 
                                    rt_window = 1/60, all_features = T, 
                                    corr_thresh = 0.9, d_thresh = 0.8 )
POS_compressedQC <- compress_clusters(POS_clusteredQC)
```

## Plotting the notame PCA workflow

```{r}
POS_PCA_2<-plot_pca(POS_compressedQC)
POS_PCA_2
```

# Second PCA and loading plot

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}
# Extracting flagged features
POS_no_flag <- drop_flagged(POS_compressedQC)

# Extracting feature table (Expression data)
POS_feature_table <- exprs(POS_no_flag)

# Extracting Phenotipic data
POS_pheno_data <- POS_no_flag@phenoData@data
```

## Preparing data

```{r}
# Transposing feature table 
POS_feat_table_pca <- t(POS_feature_table)

# Changing NA to 0 
POS_feat_table_pca[is.na(POS_feat_table_pca)]=0

# Centering and Scaling features
POS_pca_noQC <- prcomp(POS_feat_table_pca, center = T, scale. = T)
```

## Plotting PCA results

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}
POS_scores <- POS_pca_noQC$x %>%         # Get PC coordinates
  data.frame %>%                         # Convert to data frames
  mutate(Sample_ID = rownames(.)) %>%    # Create a new column with the sample names
  left_join(POS_pheno_data)              # Adding metadata

ggplot(POS_scores, aes(PC1, PC2, shape = Group, color = Group)) +
  geom_point(size = 3) +
  guides(x=guide_axis(title = "PC 1 (43%)"), y=guide_axis(title = "PC 2 (20%)")) +
  theme_classic()

# Save plot
#ggsave('Result/LC-MS_Positive/LC_MS_Positive_Score_Plot.png', width = 5, height = 4, device='png', dpi="print")
```  

## Plotting loading results

```{r}
POS_loadings <- POS_pca_noQC$rotation %>%   # Extract loadings
  data.frame(Feature_name = rownames(.))    # New column with feat name
```

Creating a table with Feature name and Compound column

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}
POS_feat_name <- readxl::read_excel("Data/POS_Metabolites.xlsx", 1)

# Creating a new small table of the annotated compounds
POS_compouds_all <- left_join(POS_feat_name, POS_loadings)

# Plotting results
ggplot(POS_loadings, aes(PC1, PC2)) + 
  geom_point(alpha = 0.1) +
  theme_classic() + 
  geom_point(data = POS_compouds_all, aes(shape = Identification_level, color = Identification_level), size = 3) +
  ggrepel::geom_label_repel(data = POS_compouds_all, aes(label = Compound),
                            box.padding = 0.3) +
  guides(x=guide_axis(title = "PC 1 (43%)"), y=guide_axis(title = "PC 2 (20%)")) +
  ggsci::scale_color_aaas()

# Save plot
#ggsave('Result/LC-MS_Positive/LC_MS_Positive_Loadings_Plot.png', width = 15, height = 9, device='png', dpi="print")
```

# Heat map plot

Loaded of identified metabolites abundance from Excel file.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}
met_hetmap <- readxl::read_excel("Data/POS_Metabolites_Hetmap.xlsx", 3)
met_hetmap <- met_hetmap[order(met_hetmap$A1),]     # sorted by column
row.names(met_hetmap) <- met_hetmap$X               # Named rows
met_hetmap_m <- data.matrix(met_hetmap)             # DataFrame
met_hetmap_m1 <- met_hetmap_m[1:15,-1]              # Delete the old rows with name
```

Logarithmic scale.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}
log_met_hetmap_m1 <- log10(met_hetmap_m1)
y <- log_met_hetmap_m1
```

## Plotting Heat map

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}
# Package installation and library loading
#if (!requireNamespace("devtools", quietly = TRUE)) {
#  install.packages("devtools")
#}
#devtools::install_github("cran/gplots")
library(gplots)

# Clusters
hr <- hclust(as.dist(1 - cor(t(y), method = "pearson")), method = "complete")
hc <- hclust(as.dist(1 - cor(y, method = "spearman")), method = "complete")
mycol <-  colorpanel(75, "blue", "white", "red")

# Heat map
heatmap.2(y,  Rowv = as.dendrogram(hr),  Colv = as.dendrogram(hc),  col = mycol,  density.info = "none",  trace = "none",  dendrogram = "both",  scale = "row",  labRow = NULL,  labCol = NULL,  margins = c(4, 12),  key = TRUE,  keysize = 1.5,  key.xlab = NA,  key.ylab = NA,)
```

## Save Heatmap plot

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}
png(filename ="Result/LC-MS_Positive/LC_MS2_Pos_HeatMap.png", width = 10, height = 6, units = "in", pointsize = 12, res=300)
heatmap.2(y, Rowv = as.dendrogram(hr), Colv = as.dendrogram(hc), col = mycol, density.info = "none", trace = "none", dendrogram = "both", scale = "row", labRow = NULL, labCol = NULL, margins = c(4, 12), key = TRUE, keysize = 0.7, key.xlab = NA, key.ylab = NA)
dev.off() 
```

Finish a record.

```{r}
finish_log()
```
